### 2011-06-17 thesis/novelty-intro.tex
Fixes here and there.
Novelty Detection by Thresholding - fixed acording to andrzej comments.

This chapter presents concepts that are used as base when adding novelty
detection capabilities on the semantic representation in \autoref{chap:novelty}.
It starts by providing the reader with a brief overview on novelty detection
techniques and related works.

After it shows that an optimal novelty detection system can be implemented by
thresholding on a order-relation defined over the inputs.
And that an equivalent order-relation can be imposed by a ratio between
conditional and unconditional probabilities.
Some discussion on how to interpret the meaning of those factors is also
presented.

@@

\section{Novelty Detection Review and Related Work}
Novelty detection, also called outlier or anomaly detection, is the
classification problem related with identification of new or unknown data
patterns that the system is not aware of~\cite{markou2003novelty}.

The ability to identify novel cases is crucial in any autonomous system
that is deployed to unknown or to uncontrolled environments, as it gives the
system the ability to detect something is not conforming to it knowledge and
therefore should proceed with caution.
It has several applications such as fault detection~\cite{tarassenko1999novelty},
intrusion detection~\cite{fan2001using},
detection of masses in mammograms~\cite{tarassenko1995novelty} or detection of
novel and useful documents~\cite{zhang2002novelty}.

Any normal classification application is also a good candidate for extending
with novelty detection, as it allows to detect samples the system has not been
trained on and therefore the classification is considered
instable~\cite{devarakota2008reliability}.
E.g.\ applied to digit-recognition~\cite{tax1998outlier}
or to detection of novel inputs on a neural network~\cite{bishop1994novelty}.

It is on the nature of unknown environments and in the infeasibility of
training the system on all possible object classes that lies the complexity of
novelty detection: it is only possible to obtain samples representing positives
examples of known cases and the lack of negative examples renders normal
classification methods unusable.

In order to become practical in real world applications, novelty detection
methods have to overcome a series of obstacles:
be able to generalize while still detecting novelty,
be resistant to noisy features,
ability to scale in feature dimension,
deal with multiple classes and performing detection efficiently as many
autonomous systems require real-time or close to real-time performance.

@@

% TODO: wuld be nice to have this one starting on a new page.. but still make
% sure it looks good and not a forced break.
% Reason to have this on a diferent page is to make it really clear novelty
% content (and not only chit-chat) starts now.
\clearpage
\section{Novelty Detection as a Threshold}
% Explain threshold approach to novelty detection
Due to the desire of robustness, a novelty detection system is often given the option
to classify non-novel samples as novel. This is implemented as a threshold that describes a trade
between error and detection: the more the system tries to increase detection, the more
prone to produce errors it becomes.

When dealing with a statistic point of view, noisy data or unstable features,
a decision to classify a sample $x$ as novel increases detection rate by
$P(novel|x)P(x)$ and introduces an associated error rate of $P(\overline{novel}|x)P(x)$.
Generalizing to a system that classifies a set of inputs $N$ as novel, the
true positive and false negative rates arise:

\begin{equation}
\label{eq:true-positive}
P(\textnormal{true positive}) = \sum_{x \in N}{P(novel|x)P(x)}
\end{equation}

\begin{equation}
\label{eq:false-positive}
P(\textnormal{false positive}) = \sum_{x \in N}{P(\overline{novel}|x)P(x)}
\end{equation}

The performance of a novelty detection system can be measured on these two
rates: the true-positive rate (\autoref{eq:true-positive}) describes the
interest in detecting as much as possible the novel cases;
and the false-positive rate (\autoref{eq:false-positive}) the interest in
having the smallest error possible.
As a given sample $x$ is included in $N$ both the true-positive and
false-positive rates never-decrease~\footnote{Note that both are described as
summations of products of probabilities (non-negative functions).} describing an
\emph{error and rejection tradeoff}~\cite{chow1970optimum} that any detection
system has to consider.

An optimal novelty detector can then be formulated as selecting a subset $N$
that achieves a given true-positive rate with the minimal false-positive
rate possible. This is equivalent to the \emph{continuous knapsack problem}
which allows a greedy solution by sorting the items with a value per weight
function.
Therefore a novelty system before classifying a sample $a$ as novel should
classify any sample $b$ satisfying \autoref{eq:knapsack} as that would achieve
a lower false-positive rate given a fixed true-positive rate.

\begin{equation}
\label{eq:knapsack}
\frac{P(\overline{novel}|b)P(b)}{P(novel|b)P(b)} < \frac{P(\overline{novel}|a)P(a)}{P(novel|a)P(a)}
\end{equation}

This relation between $a$ and $b$ can then be simplified into:

@@

Based on this, it can be said that an optimal novelty detection system is
interested in defining an order relation on all the possible inputs equivalent
to the order defined by the error rate: $P(\overline{novel}|x)$.
And any optimal detector can be described by the largest $P(\overline{novel}|x)$
accepted by it. Which is seen as threshold.

@@

Since there is only interest in maintaining the same order-relation as
$P(\overline{novel}|x)$ any constant factor can be dropped.
Leaving a ratio between a \emph{conditional} and
\emph{unconditional probability} suitable for implementing novelty detection
via thresholding.

\subsection{Conditional Probability}
The conditional probability $P(x|\overline{novel})$ describes the distribution
of the non-novel samples on the original distribution. In case the labelled data
available for the agent to learn a concept comes from the same distribution
where the system will run the correct approach is to use it as prior-information
for modelling the conditional probability.

Note that it is important for the labelled data to come from that same
distribution, otherwise it will contain some bias and lead to an incorrect
ordering. The bias, for example, can be introduced by an incorrect filtering.

>>>>

This chapter presents the main work of this thesis: a method to augment the conceptual
map with novelty detection capabilities.

The chapter starts by providing the reader with a brief overview on novelty detection
techniques and related works. Then it shows that an optimal novelty detection system
can be implemented by thresholding on a order-relation defined over the inputs.
And that an equivalent order-relation can be imposed by a ratio between
conditional and unconditional probabilities.
Some discussion on how to interpret the meaning of those factors is also
presented.

@@

\section{Novelty Detection Review and Related Work}
Novelty detection, also known as outlier or anomaly detection, is a
classification problem related to identification of new or unknown data
patterns that the system is not aware of~\cite{markou2003novelty}.

The ability to identify novel cases is crucial in any autonomous system
that is deployed to unknown or to uncontrolled environments, as it gives the
system the ability to detect that something is not conforming to its knowledge and
therefore should be treated with caution.
It has several applications such as fault detection~\cite{tarassenko1999novelty},
intrusion detection~\cite{fan2001using},
detection of masses in mammograms~\cite{tarassenko1995novelty} or detection of
novel and useful documents~\cite{zhang2002novelty}.

Any normal classification application is also a good candidate for extending
with novelty detection, as the results for samples the system has not been
trained on can be unreliable~\cite{devarakota2008reliability}.
e.g.\ applied to digit-recognition~\cite{tax1998outlier}
or to detection of novel inputs on a neural network~\cite{bishop1994novelty}.

It is in the nature of unknown environments and in the infeasibility of
the system to acquire examples of all possible classes where the complexity of
novelty detection lies: it is only possible to obtain samples representing positives
examples of known cases and the lack of negative examples renders normal
classification methods unusable.
In order to become practical in real world applications, novelty detection
methods have to overcome a series of obstacles:
be able to generalize while still detecting novelty,
be resistant to noisy features,
ability to scale in feature dimension,
deal with multiple classes and performing detection efficiently as many
autonomous systems require real-time or close to real-time performance.

@@

\section{Novelty Detection by Thresholding}
The objective of a novelty detection system is to classify a given sample $x$ as
either \emph{known}: $x$ being generated by a known class to the system, or \emph{novel}:
$x$ generated by a class unknown for the system.
Based on the ground\hyp{}truth of a sample four cases are possible:

\begin{description}
\item[true positive]  - when a system correctly   flags a sample of an unknown class as \emph{novel}.
\item[false positive] - when a system incorrectly flags a sample of a known class as \emph{novel}.
\item[true negative]  - when a system correctly   flags a sample of a known class as \emph{known}.
\item[false negative] - when a system incorrectly flags a sample of an unknown class as \emph{known}.
\end{description}

Due to noisy data, unstable features and lack of information it is impossible to develop
a method able to exactly guess the correct classification of a sample.
By modelling the outcomes with probabilities it becomes possible to handle the uncertainty
associated with each decision.
The notation $P(novel|x)$ will be used to denote the probability that the sample $x$ is in fact a
sample of an unknown class. $P(x)$ will denote the probability that sample $x$ is given to the
system to be classified. $\overline{novel}$ is also defined in such a way that
$P(\overline{novel}|x)$ measures the probability that a decision on classifying $x$ as novel is
wrong.

Additionally a decision on the novelty of a sample $x$ performed by a deterministic system
will be fully determined by the sample itself.
Which implies that any deterministic novelty detection system can be uniquely determined by the set $N$
of samples that are accepted by the classifier as \emph{novel}.
This way it is possible to define the probability of a true positive and false positive event for any
deterministic novelty detector with the following equations:

\begin{eqnarray}
P(\text{true positive})  &=& \sum_{x \in N}{P(novel|x)P(x)} \\
P(\text{false positive}) &=& \sum_{x \in N}{P(\overline{novel}|x)P(x})
\end{eqnarray}

Note that since probability functions are non-negative, it is impossible to decrease either the
true positive or the false positive probability by using a set $N' \supset N$.
This describes the base of the \emph{error and rejection tradeoff}~\cite{chow1970optimum}, which
states that a system aiming at increasing the true-positive probability will eventually increase its
false-positive error.
The true positive probability can be described as the interest in detecting as much
as possible the novel classes and the false positive probability the interest
in not incurring in too many errors.
By fixating one of those it is possible to define a novelty detection
system that achieves the maximum or minimum of the other.

This way an optimal detector can be formulated by achieving the maximum true-positive
probability without its false-positive probability increase beyond a given limit.
This is equivalent to a \emph{continuous knapsack problem} which allows a greedy
solution by sorting the items with a value per weight function.
In the case of detection system that can be defined as:

\begin{eqnarray}
value(x)  &=& P(\text{true positive}|x) \\
weight(x) &=& P(\text{false positive}|x) \\
cost(x)   &=& value(x)/cost(x) \\
          &=& \frac{P(novel|x)P(x)}{P(\overline{novel}|x)P(x)}
\end{eqnarray}

Therefore a novelty system before classifying a sample $a$ as novel should (greedily)
classify any sample $b$ with a small cost as that would achieve
an higher true positive probability given a fixed false positive one.

\begin{equation}
\label{eq:knapsack}
\frac{P(novel|b)}{P(\overline{novel}|b)} < \frac{P(novel|a)}{P(\overline{novel}|a)}
\end{equation}

This relation between $a$ and $b$ can further be simplified into:

@@

Based on this, it can be said that an optimal novelty detection system is
interested in defining an order relation on all the possible inputs equivalent
to the order defined by the function: $P(\overline{novel}|x)$.
And any optimal detector can be described by the largest $P(\overline{novel}|x)$
accepted by it. Which is seen as threshold.

@@

Since there is only interest in maintaining the same order-relation as
$P(\overline{novel}|x)$ any constant factor can be dropped.
Leaving a ratio between a \emph{conditional} and
\emph{unconditional probability} suitable for implementing novelty detection
by thresholding.

\subsection{Conditional Probability}
The conditional probability $P(x|\overline{novel})$ describes the distribution
of the samples given that they are generated by a known class.
In case the labelled data available for the agent to learn a concept comes from
the same distribution where the system will run the correct approach is to use it
as prior-information for modelling the conditional probability.

Note that it is important for the labelled data to be a filtered version
of the underlying world distribution (with all classes) that does not contains
any bias.
Otherwise that bias will lead to incorrect modelling the conditional probability
and a subsequent wrong ordering of the sample space.
